A Collection of Thoughts on the Keccak/SHA-3 Construction

### 2018-04-26 s2018-math490/thesis/thesis.tex
Changes
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

It is obvious that the order of the $\rho$ permutation is $w$: this is obvious
from the construction of $\rho$.

@@

It is obvious that the order of the $\pi$ permutation is $24$: this is obvious
from the construction of $\pi$.

@@

It is non-trivial that the order of the $\chi$ function is a constant 4.
However, this can be verified through the use of SAT to prove that $\chi^{4}$
is the identity function. Hence $\chi^{-1} = \chi^{3}$.

@@

\subsection{Generalizations of $\theta$ and $\chi$} \label{sec:p:generalizations}

@@

\subsection{Choice of Parameters and Ordering of Composition} \label{sec:p:generalizations}
% TODO Outline
%   - Pending evaluation?

@@

@@

@@

@@

>>>>

It is obvious that the order of the $\rho$ permutation is $w$: this follows
from the construction of $\rho$.

@@

It is obvious that the order of the $\pi$ permutation is $24$: this follows
from the construction of $\pi$.

@@

It is non-trivial that the order of the $\chi$ function is a constant 4.
However, this can be verified through the use of SAT to prove that $\chi^{4}$
is the identity function. Hence $\chi^{-1} = \chi^{3}$. However, due to the
construction of $\chi$, XOR does not distribute over it.

@@

% \subsection{Generalizations of $\theta$ and $\chi$} \label{sec:p:generalizations}

@@

% \subsection{Choice of Parameters and Ordering of Composition} \label{sec:p:generalizations}
% TODO Outline
%   - Pending evaluation?

@@

This section studies three approaches to extracting information about the five
permutations functions using SAT. None of these approaches can directly lead
to a collision in SHA-3, however, by combining them, possible differential
trails can be approximated. Sadly for $w \geq 4$, several of these attacks take
too long to run (even for individual permutation functions) and thus do not
emit a valid attack for full Keccak.

@@

By modeling the problem with SAT, we were able to recreate the work of prior
authors (\cite{Homsirikamol2012}, \cite{cryptoeprint:2010:285}). However, like
previous authors, our results do not scale to useful results. For $w=1$, we
have exhaustively searched the collision space; the results of this search are
reproduced below:

\begin{tabular}{c c c} \label{tab:s:1}
    $w$ & $r$ & Number of collisions \\
    1 & 1 & 1024 \\
    1 & 2 & 502 \\
    1 & 3 & 543 \\
    1 & 4 & 525 \\
    1 & 5 & 488 \\
    1 & 6 & 518 \\
    1 & 7 & 532 \\
    1 & 8 & 498 \\
    1 & 9 & 506 \\
    1 & 10 & 522 \\
    1 & 11 & 506 \\
    1 & 12 & 503 \\
    1 & 13 & 495 \\
    1 & 14 & 522 \\
    1 & 15 & 485 \\
    1 & 16 & 540 \\
    1 & 17 & 467 \\
    1 & 18 & 490 \\
    2 & 1 & $ > 2^{20}$ \\
    2 & 2 & $ > 2^{14}$ \\
    2 & 3 & $ > 74$ \\
\end{tabular}

Note that the results for $w=2$ are incomplete; the search was terminated
after one week. Further, all of these used an effective margin of 512-bits.
In our limited time, we were unable to find any useful patterns in this data.

@@

\section{Correlation Matrices} \label{sec:correlation}

Another area of study was the resistance of Keccak to correlation attacks.
That is, given some structured input (in this case, a valid block where the
remaining bits are all zeros), does SHA-3 emit any useful two bit correlations
across rounds? Here, we constructed a program to generate matricies of
correlation for all two-variable boolean functions given some state. We then
performed Monte-Carlo simulation; for $w=8$ and $r=24$, we found no useful
correlations under sufficiently large starting states ($>16$ million). Thus,
Keccak is secure from cross-round first and second order correlations; all
distinguisher attacks must thus rely on at least third-order correlations. For
$w=8$, checking all third-order correlations across 24 rounds becomes
computationally infeasible and thus was not performed, since it is not likely
to yeild useful results.

@@

For $w=1$, we reproduce a table of quantities of partial fixed points per round
below for the first 8 rounds:

\begin{tabular}{c c c c c} \label{tab:f:p:1}
    Rounds & Quantity & Input & Output \\
    1 & 512 & \texttt{FFFFFFFFFFFFFFFFFFFFFFFFF} & \texttt{TFFFFFFFFFFFFFFFFFFFFFFFF} \\
    2 & 567 & \texttt{FTFTTTTFFTTTFFTFTFFFFFFFF} & \texttt{TTTFTFFTFTTTFTFFTFFFFFFFF} \\
    3 & 527 & \texttt{FFTFFFFTFTFFFFFFFFFFFFFFF} & \texttt{TTTFTTTTFTFTFTFTFFFFFFFFF} \\
    4 & 545 & \texttt{TTTFFTTTFFTTFTTFFFFFFFFFF} & \texttt{FFFTTTTFTTFFFTFTFFFFFFFFF} \\
    5 & 488 & \texttt{FFFFFFFFTFFFTFFFFFFFFFFFF} & \texttt{TFFTFFTFFFFFFTFFTFFFFFFFF} \\
    6 & 533 & \texttt{FFTFFFFTTFFFTFFFFFFFFFFFF} & \texttt{TTTFFFFTFFFFTTFFTFFFFFFFF} \\
    7 & 510 & \texttt{TFFFTTTFTTTTFFTFFFFFFFFFF} & \texttt{FTFFFFFFFTFTTTFTTFFFFFFFF} \\
\end{tabular}

Note that the above table is only for an effective margin of 256-bits (4 bits
when $w=1$). For an effective margin of 512-bits (8 bits), we found no such
partial fixed points.

However, brute forcing a partial fixed point for larger values of $w$ and $r$
quickly grows impossible. A similar technique to the aforementioned
differential matrices could extend these techniques to larger values of $w$
and $r$. Instead of building a model to check for a specific number of
differences, instead build a model to check for a specific number of zeros in
the inputs and outputs. This could provide a reduction in number of search
paths for the SAT solver, and allow for more explicit parallelism.
